A Study on Common Bug Fix Patterns in Rust

### 2022-10-29 results.tex
comment about rq1, and rephrase
flattened everything ;((

@@

As discussed in the previous section, we used an encoding mechanism to incorporate AST information in fix-sized datapoints which then could be used in the DBSCAN clustering algorithm. We used heuristics (weighting scheme) while encoding our datapoints to make the most important program elements stand out between all the non terminals seen in the AST. Here we want to evaluate the effectiveness of our encoding.

The general idea for our evaluation is whether the most important program elements has the high values in our encoding. The analyzer compares the actual change in the code and compares it to the visual representation of its respective data point. We used circle pack figures to visualize each data point. In these figures, the radius of the circle of each non terminal shows its importance in our encoding (Fig ref).

@@

We carried out two manual experiments for evaluating our encoding mechanism. In the first experiment, the analyzer investigates whether the most important element has the largest radius or not. In the second experiment, the analyzer investigates whether the most important element is included in three largest circles or not. The analyzer should give a binary answer to these questions. Both authors carry out these experiments independently. At last, we report the results as the portion of the changes which both authors agreed upon its encoding correctness. The final results are TBD\% and TBD\% correctness in the first and second experiments, respectively, which shows the effectiveness of our encoding mechanism.

\subsection{Common Bug Fix Patterns (RQ2)}

\subsection{Attributes}

Attributes in Rust make it possible for the developers to add metadata to program elements. These attributes can be added to structs, struct fields, enum variants, match expression arms, etc. There are two types of attributes: Inner Attributes and Outer Attributes. Inner Attributes apply to the program element encapsulating them, while Outer Attributes apply to the program element coming after the attribute. \\

\noindent \textit{ \textbf{G.attr.struct.} Modifying the attributes of structs}

\begin{lstlisting}[language=Rust, style=colouredRust, label={l3}]
// G.attr.struct.add: Adding attributes to a struct
++ #[derive(Debug, Clone)]
pub struct Call {
/// body 
}

//G.attr.struct.drop: Removing attributes from a struct
-- #[derive(Debug, Clone)]
pub struct Call {
/// body 
}

// G.attr.struct.change: Changing an existing attribute of a struct
-- #[derive(Debug)]
++ #[derive(Debug, Clone)]
pub struct Call {
/// body 
}
\end{lstlisting}

\noindent \textbf{Description:} In this pattern, the attribute set of a struct changes. This modification can happen in three different ways; (G1.1) Adding attributes to the struct, (G1.2) Removing attributes from the struct, (G1.3) Changing the content of an existing attribute. The attributes are Outer Attributes as we can find struct related non-terminals within the ASTDiff of the instances of this pattern. \\

\noindent \textit{ \textbf{G.attr.struct.field} Modifying the attributes of struct fields}

\begin{lstlisting}[language=Rust, style=colouredRust, label={l3}]
// G.attr.struct.field.add: Adding attributes to struct fields
#[derive(Serialize)]
pub struct Retain {
    ++ #[serde(skip_serializing_if = "is_empty")]
    pub attributes: Option<Attributes>,
}

// G.attr.struct.field.drop: Removing attributes from struct fields
#[derive(Serialize)]
pub struct Retain {
    -- #[serde(skip_serializing_if = "Option::is_none")]
    pub attributes: Option<Attributes>,
}

// G.attr.struct.field.change: Changing attributes of struct fields
#[derive(Serialize)]
pub struct Retain {
    -- #[serde(skip_serializing_if = "Option::is_none")]
    ++ #[serde(skip_serializing_if = "is_empty")]
    pub attributes: Option<Attributes>,
}
\end{lstlisting}

\noindent \textbf{Description:} In this pattern, the attribute set of a struct field changes. As with G1, this modification also happens in three different ways. Using attributes for struct fields, apart from being able to store different meta-information for different struct fields, for some use cases, the programmers can control the action they want to apply on different fields.

\subsection{Struct} 
\noindent \textit{ \textbf{G.struct.field} Adding/Removing a struct field}

\begin{lstlisting}[language=Rust, style=colouredRust, label={l3}]
// G.struct.field.add: Adding a new field to a struct
struct InnerListeners {
    pending: Mutex<Vec<Pending>>,
    <@\textcolor{red}{--}@> queue_object_name: Uuid,
}

@@

\noindent \textbf{Description:} A set of new fields are add to or removed from an existing struct. The addition of a new field happens when the developer need to store a new piece of data so they can implement a desired behaviour. The removal of a field  happens when the developer realizes that the field is expendable with respect to the behaviour they want to implement. \\

\noindent \textit{ \textbf{G.struct.field.pub} Modifying the access modifiers of struct fields}

\begin{lstlisting}[language=Rust, style=colouredRust, label={l3}]

// G.struct.field.pub.add: Adding pub to a field
pub struct HtmlBlock {
    -- content: BlockContent,
    ++ pub content: BlockContent,
    brace: token::Brace,
}

// G.struct.field.pub.drop: Removing pub from a field
pub struct Languages {
--  pub named: HashMap<LanguageId, Arc<LanguageDefinition>>,
++  named: HashMap<LanguageId, Arc<LanguageDefinition>>,
    extensions: HashMap<String, Arc<LanguageDefinition>>,
}

@@

\noindent \textbf{Description:} Adding pub to a struct field makes it possible to access that field within external modules. In the instances of this cluster, adding pub access modifier happened in a bug fixing change, where the developer needed to access a field which had not been granted public access. On the other hand, the public access might be revoked if the developer realizes there is no external access to the field.

\subsection{Option}
\vspace{3mm}

\noindent \textit{ \textbf{G.option} Changing field type T to T$\langle$Option$\rangle$}

\begin{lstlisting}[language=Rust, style=colouredRust, label={l3}]

#[ast_node("MediaRule")]
pub struct MediaRule {
    pub span: Span,
    -- pub media: MediaQueryList,
    ++ pub media: Option<MediaQueryList>,
    pub rules: Vec<Rule>,
}

@@

\noindent \textbf{Description:} Rust is a null safe language, meaning that object references cannot assume null values. If we want void values for our object, we would have to use Option type. Option is essentially an enum with two variants, Some which indicates that our object has an value, and None shows the absence of any value. In this pattern, the developer changes the type of a struct field to an Option of its type. This makes it possible for the developer to modify the program in a way that it covers the scenarios in which the the struct field has no value.

For instance, in the Rust project swc-project/swc (A fast TypeScript/JavaScript compiler), the developer needed to modify the CSS parser source code in a way that it accounts for empty @media queries \footnote{\url{https://github.com/swc-project/swc/commit/75a14f98b7370226115ee24eec6eb8c802bd4837}}, so, as shown in the code snippet above, they changed type MediaQueryList to Option<MediaQueryList>, and modified the other parts of the source code, accordingly.

\subsection{Types}
\vspace{3mm}

\noindent \textit{ \textbf{G.type.field} Changing a struct field type}

\begin{lstlisting}[language=Rust, style=colouredRust, label={l3}]
pub struct Manifest {
  -- pub features: Vec<String>,
  ++ pub features: HashSet<String>,
}

@@

\noindent \textbf{Description:} The type of a struct field changes. This is a common change pattern in all statically-typed languages and can happen for various reasons. These reasons can be a step for implementing a new feature or rectifying the behaviour of the program. Also it can happen for refactoring purposes or performance enhancement. \\

\noindent \textit{ \textbf{G.type.generic} Changing a generic type}

\begin{lstlisting}[language=Rust, style=colouredRust, label={l3}]
#[derive(Debug, Clone, PartialEq)]
pub struct Anchor {
    -- point: Point<usize>,
    ++ point: Point<isize>,
    side: Side,
}

@@

\noindent \textbf{Description:} This a change of the generic (parametric) type of a struct field. Similar to G7, this change can also happen for the mentioned reasons. \\

\noindent \textit{ \textbf{G.type.enum.variant} Change in enum variant value type}

\begin{lstlisting}[language=Rust, style=colouredRust, label={l3}]
#[derive(Debug, Clone, PartialEq, Eq)]
pub enum ResolvedDependency {
  Resolved(ModuleSpecifier),
  -- Err(String),
  ++ Err(ResolvedDependencyErr),
}

@@

\noindent \textbf{Description:} In Rust, we can specify the type of the value that an enum variant can store. Using this feature, we can avoid using a struct alongside our enum variants to specify the value type. This change pattern is associated with a change in enum variant types. \\

\noindent \textit{ \textbf{G.type.tuple} Changing the type inside a tuple}

\begin{lstlisting}[language=Rust, style=colouredRust, label={l3}]
#[allow(dead_code)]
pub struct CoreState {
    // Other fields
    -- pending_views: Vec<ViewId>,
    ++ pending_views: Vec<(ViewId, Table)>,
    peer: Client
}
\end{lstlisting}

\noindent \textbf{Description:} In this pattern, an extra type is added to an existing type, making it a tuple of multiple types. Alternatively, the change might be dropping types from the tuple and making it a tuple with fewer number of arguments.

\subsection{Traits}
\noindent \textit{ \textbf{G.trait.bound} TraitBound Change}

\begin{lstlisting}[language=Rust, style=colouredRust, label={l3}]
-- pub struct EventLoop<T: tty::EventedReadWrite> {
++ pub struct EventLoop<T: tty::EventedPty> {
    poll: mio::Poll,
    // Other fields
}

@@

\noindent \textbf{Description:} The trait bounds are the functionalities that we require from our parametric types. This change pattern related to modification in trait bounds. This can happen both for bug fix reasons or just refactoring purposes.

\subsection{BC-Related Bug Fix Patterns (RQ3)}

@@

\noindent \textit{ \textbf{BC.clone.ref} Dropping clone and adding borrowing}

\begin{lstlisting}[language=Rust, style=colouredRust, label={l3}]
-- let field_id = schema.get_or_create(attribute.clone())?;
++ let field_id = schema.get_or_create(&attribute)?;
\end{lstlisting}

\noindent \textbf{Description:} Unlike the previous pattern, this pattern is not detected by Clippy. The change happens when the programmers realizes that the cloning of a variable is unnecessary, however, we also want to preserve the ownership of the variable within the current scope. That is why in this pattern a clone is turned into a borrowing.

Similar to the previous pattern, this does not change the program behaviour and is merely for performance purposes. It was the case in one of the commits of the project tauri-apps/tauri, where the clone to borrowing significantly optimized CPU usage. They discussed that the cloning of some types such as HashMap was expensive. \\

\noindent \textit{ \textbf{BC.clone.drop} Removing clone}

\begin{lstlisting}[language=Rust, style=colouredRust, label={l3}]
start_plugin_process(
--  manifest.clone(),
++  manifest,
    self.next_plugin_id(),
    self.self_ref.as_ref().unwrap().clone(),
);
\end{lstlisting}

\noindent \textbf{Description:} This pattern is referred to as redundant clone in Clippy Lints. Clippy detects whether a variable clone is necessary and could we have used the variable itself or not. This change does not modify the program behaviour, however it can boost the performance significantly if we repeatedly cloning large variables.

@@

\noindent \textit{ \textbf{BC.ref.add} Adding Borrowing}

\begin{lstlisting}[language=Rust, style=colouredRust, label={l3}]
let repo = replace(&mut *contents, Processing).inner_repo();
-- let statuses = repo_to_statuses(repo, &self.workdir);
++ let statuses = repo_to_statuses(&repo, &self.workdir);
\end{lstlisting}

\noindent \textbf{Description:} This change happens when the developer needs to modify the program so it reaches the desired behaviour. The modification, however, requires access to an object whose ownership has moved to another scope and never returned. That is why we must use borrowing to preserve ownership within our scope, after the function execution finishes.

To give an example, in starship/starship after one of the commits \footnote{\url{https://github.com/starship/starship/commit/56d475578ea508631275772127f49a6949fea6b0}{git_branches.rs}}, the developer borrows the value of $repo\_dir$ to keep the ownership of the object in the current scope. Now this object can then be used in another function invocation ($remove\_dir\_all(repo\_dir)$) which fixes a bug. \\

\noindent \textit{ \textbf{BC.ref.drop} Removing Borrowing}

\begin{lstlisting}[language=Rust, style=colouredRust, label={l3}]
enums.iter().for_each(|e| {
    let mut enum_template = EnumTemplate::new();
    -- enum_template.set_message_enum(&e);
    ++ enum_template.set_message_enum(e);
    // Other statements
});
\end{lstlisting}

\noindent \textbf{Description:} In the pattern, an unnecessary borrowing is removed. Clippy can detect the needless borrows, and also can automatically remove them. This does not change the behaviour of the program, however, it does improve the performance, as the borrow checker need not to return the ownership after the execution of the function. \\

\noindent \textit{ \textbf{BC.deref.to\_string()} Modification in dereferencing a string prior to calling to\_string()}

\begin{lstlisting}[language=Rust, style=colouredRust, label={l3}]
for p in wixobjs {
--  args.push(p.to_string());
++  args.push((*p).to_string());
}

let displayed_attributes = schema.clone().map(|s| {
    s.displayed_name()
        .iter()
--      .map(|s| (*s).to_string())
++      .map(|s| s.to_string())
        .collect::<HashSet<String>>()
});
\end{lstlisting}

\noindent \textbf{Description:} This is a change that in all the instances has been detected by Clippy. The change applies on an object prior to calling $to\_string()$ on that object. The change might add or remove a dereference, as shown above. In the addition case, the reason behind the change is that the object a reference type of $T$, while $T$ directly implements $to\_string()$. It is faster and more performance efficient to dereference the object and then call $to\_string()$ on it. The deletion case indicates a redundant dereference, thus it is removed.

@@

\noindent \textit{ \textbf{BC.mut.add} Adding mutability}

\begin{lstlisting}[language=Rust, style=colouredRust, label={l3}]
parser::Parser::new(args)
        .parse_module()
--      .map_err(|e| {
++      .map_err(|mut e| {
            e.emit();
            ()
        })?
\end{lstlisting}

\noindent \textbf{Description:} This change happens when the programmer needs to mutate a variable which is immutable, hence they need to add the mut keyword. In the observed instances, this pattern was not seen in refactoring changes, the modification happened alongside a change in program behaviour.

\noindent \textit{ \textbf{BC.mut.drop} Dropping mutability}

\begin{lstlisting}[language=Rust, style=colouredRust, label={l3}]
-- let mut tx = tx.clone();
++ let tx = tx.clone();
\end{lstlisting}

\noindent \textbf{Description:} Dropping the mut keyword before a variable, as we do not mutate the variable. This redundant mutability is reported by Clippy and can also be removed through this tool.

@@

\noindent \textit{ \textbf{BC.vec.slice} Changing a Vec reference to Slice}

\begin{lstlisting}[language=Rust, style=colouredRust, label={l3}]
pub fn expand_delimited_square(
--  children: &Vec<TokenNode>,
++  children: &[TokenNode],
) -> Result<hir::Expression, ParseError> {
    // body
}
\end{lstlisting}
\noindent \textbf{Description:} In struct field definition or the types of the function formal parameters, type $\&Vec\<T\>$ changes to $\&[T]$. Clippy can detect the opportunity to apply this pattern. The reason is reducing type complication, as slice types in the form of $\&[T]$ or $\&str$ are usually sufficient for most use cases.

\subsubsection{Lifetime}
\noindent \textit{ \textbf{BC.lifetime.static.drop} Dropping static lifetime}

\begin{lstlisting}[language=Rust, style=colouredRust, label={l3}]
-- const QUEUE_SIZES: &'static [u16] = &[QUEUE_SIZE];
++ const QUEUE_SIZES: &[u16] = &[QUEUE_SIZE];

@@

\noindent \textbf{Description:} Removing the static keyword from a const variable declaration. This change can be applied using Clippy. If the presence of static is not required, it is better that we omit it as keeping it might create complicate types in our program.

>>>>

In this section, we present the clusters of fix patterns that we found in our data set. \RQBC{} is about the borrow-checker related patterns, while \RQG{} is about general patterns. However, our technique fundamentally relies on our data modelling approach successfully capturing the most important aspects of program changes. We thus first evaluate the data modelling approach as our RQ1, before continuing with \RQBC{} and \RQG{}.

@@

\textbf{Does our data modelling approach capture the most important aspect of program changes?}

As discussed in the previous section, we used an embedding mechanism to incorporate AST information in fixed-sized datapoints which then could be used in the DBSCAN clustering algorithm. We used heuristics (weighting scheme) while creating our datapoints to make the most important program elements stand out between all the non terminals seen in the AST (alongside some BC-related elements). Here we want to evaluate the effectiveness of our code embedding.

The general idea for our evaluation is whether the most important program elements get the high values in our embedding. The analyzer compares the actual change in the code and compares it to the visual representation of its respective data point. We used circle pack figures to visualize each data point~\cite{collins2003circle}. For this purpose, we used the packcircles python library\footnote{\url{https://github.com/mhtchan/packcircles}}. In these figures, the radius of the circle of each non terminal shows its importance in our embedding (similar to Figure~\ref{fig:essence}).

@@

We carried out two manual experiments for evaluating our embedding mechanism. In the first experiment, the analyzer investigates whether the most important element has the largest radius or not. In the second experiment, the analyzer investigates whether the most important element is included in three largest circles or not. The analyzer should give a binary answer to these questions. Both authors carried out these experiments independently. At last, we report the results as the portion of the changes which both authors agreed upon its embedding correctness. The final results are 54\% and 80\% correctness in the first and second experiments, respectively.

% no, you have to compute the intersection
% when is our approach effective and when is it not?
% more detailed stats would probably be good too

\subsection{\label{sec:common_patterns}Common Bug Fix Patterns (RQ2)}

\textbf{What are the general fix patterns in Rust, and how often do they apply?}

We collected 60583 datapoints for general fix patterns using a server with Xeon Gold 5120 Processor (Q3-2017, 14-core 2.2 to 3.2GHz) and the data collection took around 26 hours to complete for both $D_g$ and $D_b$. We used $\epsilon=0.0001$ and $Z=5$ as our DBSCAN parameters and that resulted in 577 clusters. After manual analysis as described in~\ref{sec:manual_analysis_cluster_selection} we ended up with 10 general patterns which we introduce here:

\begin{tabular}{l|l|r|r}
\textbf{Catgegory} & \textbf{ID} & \textbf{\# Instances} & \textbf{\# Projects} \\
\hline
\multirow{2}{*}{Attributes} & G.attr.struct & 172 & 16                                  \\
& G.attr.struct.field & 62 & 16                     \\\cline{1-4}
\multirow{2}{*}{Struct} & G.struct.field  & 179 & 17\\
& G.struct.field.pub & 23 & 12   
\\\cline{1-4}
Option & G.option & 22 & 12                    \\\cline{1-4}
\multirow{4}{*}{Types} & G.type.field  & 72 & 14 \\
& G.type.generic & 46 & 14\\ & G.type.enum.variant  & 39 & 12 \\
& G.type.tuple & 6 & 6                                         \\\cline{1-4}
\multirow{1}{*}{Traits} & G.trait.bound & 6 & 5 \\ 
\end{tabular}
\caption{\label{table:general}Bug Fix Pattern Categorization (General Patterns)}
\end{table}

\subsubsection{Attributes}

Attributes in Rust enable developers to add metadata to program elements. Developers can add attributes to structs, struct fields, enum variants, match expression arms, etc. There are two types of attributes: Inner Attributes and Outer Attributes. Inner Attributes apply to the program element encapsulating them, while Outer Attributes apply to the program element coming after the attribute. \\

\noindent\textit{\label{sec:G.attr.struct}\textbf{G.attr.struct} Modifying the attributes of structs}

\begin{lstlisting}[language=Rust, style=colouredRust]
// G.attr.struct.add: Adding attributes to a struct
<@\textcolor{green}{++}@> #[derive(Debug, Clone)]
pub struct Call {
/// body 
}

//G.attr.struct.drop: Removing attributes from a struct
<@\textcolor{red}{--}@> #[derive(Debug, Clone)]
pub struct Call {
/// body 
}

// G.attr.struct.change: Changing an existing attribute of a struct
<@\textcolor{red}{--}@> #[derive(Debug)]
<@\textcolor{green}{++}@> #[derive(Debug, Clone)]
pub struct Call {
/// body 
}
\end{lstlisting}

\noindent\textbf{Description:} In this pattern, the attribute set of a struct changes. This modification can happen in three different ways: (G.attr.struct.add) adding attributes to the struct; (G.attr.struct.drop) removing attributes from the struct; and (G.attr.struct.change) changing the content of an existing attribute. These attributes are Outer Attributes as we can find struct related non-terminals within the ASTDiff of the instances of this pattern. \\

\noindent\textit{\textbf{G.attr.struct.field} Modifying the attributes of struct fields}

\begin{lstlisting}[language=Rust, style=colouredRust]
// G.attr.struct.field.add: Adding attributes to struct fields
#[derive(Serialize)]
pub struct Retain {
    <@\textcolor{green}{++}@> #[serde(skip_serializing_if = "is_empty")]
    pub attributes: Option<Attributes>,
}

// G.attr.struct.field.drop: Removing attributes from struct fields
#[derive(Serialize)]
pub struct Retain {
    <@\textcolor{red}{--}@> #[serde(skip_serializing_if = "Option::is_none")]
    pub attributes: Option<Attributes>,
}

// G.attr.struct.field.change: Changing attributes of struct fields
#[derive(Serialize)]
pub struct Retain {
    <@\textcolor{red}{--}@> #[serde(skip_serializing_if = "Option::is_none")]
    <@\textcolor{green}{++}@> #[serde(skip_serializing_if = "is_empty")]
    pub attributes: Option<Attributes>,
}
\end{lstlisting}

\noindent\textbf{Description:} In this pattern, the attribute set of a struct field changes. As with \textit{\textbf{G.attr.struct}}, this modification also happens in three different ways. Using attributes for struct fields, apart from being able to store different meta-information for different struct fields, for some use cases, the programmers can control the action they want to apply on different fields.
% this last sentence immediately prior to here has too many moving parts

\subsubsection{Struct} 
\noindent\textit{\textbf{G.struct.field} Adding/Removing a struct field}

\begin{lstlisting}[language=Rust, style=colouredRust]
// G.struct.field.add: Adding a new field to a struct
struct InnerListeners {
    pending: Mutex<Vec<Pending>>,
    <@\textcolor{red}{--}@> queue_object_name: Uuid,
}

@@

\noindent\textbf{Description:} A set of new fields are added to or removed from an existing struct. A developer adds a new field to store a new piece of data in a struct. A developer removes a field when the developer realizes that the field is not required with respect to the behaviour they want to implement (possibly because they have moved it elsewhere). \\

\noindent\textit{\textbf{G.struct.field.pub} Modifying the access modifiers of struct fields}

\begin{lstlisting}[language=Rust, style=colouredRust]

// G.struct.field.pub.add: Adding pub to a field
pub struct HtmlBlock {
    <@\textcolor{red}{--}@> content: BlockContent,
    <@\textcolor{green}{++}@> pub content: BlockContent,
    brace: token::Brace,
}

// G.struct.field.pub.drop: Removing pub from a field
pub struct Languages {
<@\textcolor{red}{--}@>  pub named: HashMap<LanguageId, Arc<LanguageDefinition>>,
<@\textcolor{green}{++}@>  named: HashMap<LanguageId, Arc<LanguageDefinition>>,
    extensions: HashMap<String, Arc<LanguageDefinition>>,
}

@@

\noindent\textbf{Description:} Adding \texttt{pub} to a struct field makes it possible to access that field from external modules. In the instances of this cluster, adding the \texttt{pub} access modifier happened in a bug fixing change, where the developer needed to access a field which had not been marked for public access. On the other hand, the public access might be revoked if the developer realizes there are no external accesses to the field, and that such accesses are not desirable, e.g. for encapsulation purposes.

\subsubsection{Option}
\vspace{3mm}

\noindent\textit{\textbf{G.option} Changing field type T to T$\langle$Option$\rangle$}

\begin{lstlisting}[language=Rust, style=colouredRust]

#[ast_node("MediaRule")]
pub struct MediaRule {
    pub span: Span,
    <@\textcolor{red}{--}@> pub media: MediaQueryList,
    <@\textcolor{green}{++}@> pub media: Option<MediaQueryList>,
    pub rules: Vec<Rule>,
}

@@

\noindent\textbf{Description:} Rust is a null-safe language, meaning that object references cannot take on null values. If we want a variable to contain null, we would have to use the \texttt{Option} type. \texttt{Option} is essentially an \texttt{enum} with two variants, \texttt{Some}, which indicates that our variable has a value, and \texttt{None}, which indicates the absence of any value. In this change pattern, the developer changes the type of a struct field to an \texttt{Option} of that type, hence enabling them to store no value in that field.

For instance, in the Rust project swc-project/swc (a fast TypeScript/JavaScript compiler), the developer needed to modify the CSS parser source code in a way that it accounts for empty @media queries \footnote{\url{https://github.com/swc-project/swc/commit/75a14f98b7370226115ee24eec6eb8c802bd4837}}. So, as shown in the code snippet above, they changed type \texttt{MediaQueryList} to \verb+Option<MediaQueryList>+, and modified the other parts of the source code, accordingly.

\subsubsection{Types}
\vspace{3mm}

\noindent\textit{\label{sec:G.type.field}\textbf{G.type.field} Changing a struct field type}

\begin{lstlisting}[language=Rust, style=colouredRust]
pub struct Manifest {
  <@\textcolor{red}{--}@> pub features: Vec<String>,
  <@\textcolor{green}{++}@> pub features: HashSet<String>,
}

@@

\noindent\textbf{Description:} The type of a struct field changes. This is a common change pattern in all statically-typed languages and can happen for various reasons. For instance, the developer may be implementing a new feature or fixing the behaviour of the program. Also it can happen for refactoring purposes or performance enhancement. \\
% maybe discuss the specific example a bit?

\noindent\textit{\textbf{G.type.generic} Changing a generic type parameter}

\begin{lstlisting}[language=Rust, style=colouredRust]
#[derive(Debug, Clone, PartialEq)]
pub struct Anchor {
    <@\textcolor{red}{--}@> point: Point<usize>,
    <@\textcolor{green}{++}@> point: Point<isize>,
    side: Side,
}

@@

\noindent\textbf{Description:} This a change of the type parameter of a struct field. As with \textit{\textbf{G.type.field}}, this change can also happen for the mentioned reasons. \\

\noindent\textit{\textbf{G.type.enum.variant} Change in enum variant value type}

\begin{lstlisting}[language=Rust, style=colouredRust]
#[derive(Debug, Clone, PartialEq, Eq)]
pub enum ResolvedDependency {
  Resolved(ModuleSpecifier),
  <@\textcolor{red}{--}@> Err(String),
  <@\textcolor{green}{++}@> Err(ResolvedDependencyErr),
}

@@

\noindent\textbf{Description:} In Rust, developers can specify the type of the value that an enum variant can store. Using this feature, they can avoid having to use a struct along with enum variants. This change pattern is associated with a change in enum variant types. \\

\noindent\textit{\textbf{G.type.tuple} Changing the type inside a tuple}

\begin{lstlisting}[language=Rust, style=colouredRust]
#[allow(dead_code)]
pub struct CoreState {
    // Other fields
    <@\textcolor{red}{--}@> pending_views: Vec<ViewId>,
    <@\textcolor{green}{++}@> pending_views: Vec<(ViewId, Table)>,
    peer: Client
}
\end{lstlisting}

\noindent\textbf{Description:} In this pattern, an extra type is added to an existing type, making it a tuple of multiple types. Alternatively, the change might be dropping types from the tuple and making it a smaller-arity tuple.

\subsubsection{Traits}
\noindent\textit{\textbf{G.trait.bound} TraitBound Change}

\begin{lstlisting}[language=Rust, style=colouredRust]
<@\textcolor{red}{--}@> pub struct EventLoop<T: tty::EventedReadWrite> {
<@\textcolor{green}{++}@> pub struct EventLoop<T: tty::EventedPty> {
    poll: mio::Poll,
    // Other fields
}

@@

\noindent\textbf{Description:} The trait bounds are the functionalities that we require from our parametric types. This change pattern relates to modification in trait bounds. This can happen both for bug fix reasons or just refactoring purposes.

\subsection{\label{sec:bc_patterns}BC-Related Bug Fix Patterns (RQ3)}

\textbf{What are borrow-checker related fix patterns in Rust, and how often do they apply?}

We collected 27143 datapoints for borrow-checker related fix patterns. We used $\epsilon=0.001$ and $Z=5$ as our DBSCAN parameters and that resulted in 102 clusters. After manual analysis as described in~\ref{sec:manual_analysis_cluster_selection} we ended up with 9 borrow-checker related fix patterns which we introduce here:

\begin{table}[]
\begin{tabular}{l|l|r|r}
\textbf{Catgegory} & \textbf{ID} & \textbf{\# Instances} & \textbf{\# Projects} \\
\hline
\multirow{2}{*}{Clone}                                         & BC.clone.drop & 83 & 10 \\
& BC.clone.ref & 4 & 3   
\\\cline{1-4}
\multirow{3}{*}{Ref and Deref} & BC.ref.add & 17 & 7 \\
& BC.ref.drop & 145 & 14 \\
& BC.deref.to\_string  & 7 & 4                                            \\\cline{1-4}
\multirow{2}{*}{Mut} & BC.mut.add  & 11 & 6 \\
& BC.mut.drop  & 42 & 15                               \\\cline{1-4}
Vector & BC.vec.slice  & 8 & 6                                    \\\cline{1-4}
\multirow{1}{*}{Lifetime}      
& BC.lifetime.static  & 8 & 6
\\
\end{tabular}
\caption{\label{table:bc}Bug Fix Pattern Categorization (BC-Related Patterns)}
\end{table}

@@

\noindent\textit{\textbf{BC.clone.drop} Removing clone}

\begin{lstlisting}[language=Rust, style=colouredRust]
start_plugin_process(
<@\textcolor{red}{--}@>  manifest.clone(),
<@\textcolor{green}{++}@>  manifest,
    self.next_plugin_id(),
    self.self_ref.as_ref().unwrap().clone(),
);
\end{lstlisting}

\noindent\textbf{Description:} This pattern is referred to as redundant clone in Clippy Lints. Clippy\footnote{\url{https://github.com/rust-lang/rust-clippy}} is a linter tool for Rust that catches common mistakes. There are more than 500 lints included in Clippy. With regards to this pattern, Clippy detects whether a variable clone is necessary and could we have used the variable itself or not. This change does not modify the program behaviour, however it can boost the performance significantly if we repeatedly cloning large variables.

\noindent\textit{\textbf{BC.clone.ref} Dropping clone and adding borrowing}

\begin{lstlisting}[language=Rust, style=colouredRust]
<@\textcolor{red}{--}@> let field_id = schema.get_or_create(attribute.clone())?;
<@\textcolor{green}{++}@> let field_id = schema.get_or_create(&attribute)?;
\end{lstlisting}

\noindent\textbf{Description:} Unlike the previous pattern, this pattern is not detected by Clippy. The change happens when the programmers realizes that the cloning of a variable is unnecessary, however, we also want to preserve the ownership of the variable within the current scope. That is why in this pattern a clone is turned into a borrowing.

Similar to the previous pattern, this does not change the program behaviour and is merely for performance purposes. It was the case in one of the commits~\footnote{\url{https://github.com/tauri-apps/tauri/commit/a280ee90af0749ce18d6d0b00939b06473717bc9}} of the project tauri-apps/tauri, where the clone to borrowing significantly optimized CPU usage. They discussed that the cloning of some types such as HashMap was expensive. \\

@@

\noindent\textit{\textbf{BC.ref.add} Adding Borrowing}

\begin{lstlisting}[language=Rust, style=colouredRust]
let repo = replace(&mut *contents, Processing).inner_repo();
<@\textcolor{red}{--}@> let statuses = repo_to_statuses(repo, &self.workdir);
<@\textcolor{green}{++}@> let statuses = repo_to_statuses(&repo, &self.workdir);
\end{lstlisting}

\noindent\textbf{Description:} This change happens when the developer needs to modify the program so it reaches the desired behaviour. The modification, however, requires access to an object whose ownership has moved to another scope and never returned. That is why we must use borrowing to preserve ownership within our scope, after the function execution finishes.

To give an example, in starship/starship after one of the commits \footnote{\url{https://github.com/starship/starship/commit/56d475578ea508631275772127f49a6949fea6b0}}, the developer borrows the value of \verb+repo_dir+ to keep the ownership of the object in the current scope. Now this object can then be used in another function invocation (\verb+remove_dir_all(repo_dir)+) which fixes a bug. \\

\noindent\textit{\textbf{BC.ref.drop} Removing Borrowing}

\begin{lstlisting}[language=Rust, style=colouredRust]
enums.iter().for_each(|e| {
    let mut enum_template = EnumTemplate::new();
    <@\textcolor{red}{--}@> enum_template.set_message_enum(&e);
    <@\textcolor{green}{++}@> enum_template.set_message_enum(e);
    // Other statements
});
\end{lstlisting}

\noindent\textbf{Description:} In the pattern, an unnecessary borrowing is removed. Clippy can detect the needless borrows, and also can automatically remove them. This does not change the behaviour of the program, however, it does improve the performance, as the borrow checker need not to return the ownership after the execution of the function. \\

\noindent\textit{\textbf{BC.deref.to\_string()} Modification in dereferencing a string prior to calling to\_string()}

\begin{lstlisting}[language=Rust, style=colouredRust]
for p in wixobjs {
<@\textcolor{red}{--}@>  args.push(p.to_string());
<@\textcolor{green}{++}@>  args.push((*p).to_string());
}

let displayed_attributes = schema.clone().map(|s| {
    s.displayed_name()
        .iter()
<@\textcolor{red}{--}@>      .map(|s| (*s).to_string())
<@\textcolor{green}{++}@>      .map(|s| s.to_string())
        .collect::<HashSet<String>>()
});
\end{lstlisting}

\noindent\textbf{Description:} This is a change that in all the instances has been detected by Clippy. The change applies on an object prior to calling \verb+to_string()+ on that object. The change might add or remove a dereference, as shown above. In the addition case, the reason behind the change is that the object a reference type of \verb+T+, while \verb+T+ directly implements \verb+to_string()+. It is faster and more performance efficient to dereference the object and then call \verb+to_string()+ on it. The deletion case indicates a redundant dereference, thus it is removed.

@@

\noindent\textit{\textbf{BC.mut.add} Adding mutability}

\begin{lstlisting}[language=Rust, style=colouredRust]
parser::Parser::new(args)
        .parse_module()
<@\textcolor{red}{--}@>      .map_err(|e| {
<@\textcolor{green}{++}@>      .map_err(|mut e| {
            e.emit();
            ()
        })?
\end{lstlisting}

\noindent\textbf{Description:} This change happens when the programmer needs to mutate a variable which is immutable, hence they need to add the mut keyword. In the observed instances, this pattern was not seen in refactoring changes, the modification happened alongside a change in program behaviour. \\

\noindent\textit{\textbf{BC.mut.drop} Dropping mutability}

\begin{lstlisting}[language=Rust, style=colouredRust]
<@\textcolor{red}{--}@> let mut tx = tx.clone();
<@\textcolor{green}{++}@> let tx = tx.clone();
\end{lstlisting}

\noindent\textbf{Description:} Dropping the mut keyword before a variable, as we do not mutate the variable. This redundant mutability is reported by Clippy and can also be removed through this tool.

@@

\noindent\textit{\textbf{BC.vec.slice} Changing a Vec reference to Slice}

\begin{lstlisting}[language=Rust, style=colouredRust]
pub fn expand_delimited_square(
<@\textcolor{red}{--}@>  children: &Vec<TokenNode>,
<@\textcolor{green}{++}@>  children: &[TokenNode],
) -> Result<hir::Expression, ParseError> {
    // body
}
\end{lstlisting}
\noindent\textbf{Description:} In struct field definition or the types of the function formal parameters, type \verb+&Vec<T>+ changes to \verb+&[T]+. Clippy can detect the opportunity to apply this pattern. The reason is reducing type complication, as slice types in the form of \verb+&[T]+ or \verb+&str+ are usually sufficient for most use cases.

\subsubsection{Lifetime}
\noindent\textit{\textbf{BC.lifetime.static.drop} Dropping static lifetime}

\begin{lstlisting}[language=Rust, style=colouredRust]
<@\textcolor{red}{--}@> const QUEUE_SIZES: &'static [u16] = &[QUEUE_SIZE];
<@\textcolor{green}{++}@> const QUEUE_SIZES: &[u16] = &[QUEUE_SIZE];

@@

\noindent\textbf{Description:} Removing the static keyword from a const variable declaration. This change can be applied using Clippy. If the presence of static is not required, it is better that we omit it as keeping it might create complicate types in our program.
